The Kapowski Chronicles: A Complete, Volumetric Cortical Thickness Open Source Pipeline with Evaluation on Public Data

### 2013-07-07 KapowskiChronicles/intro.tex
ENH:  Adding evaluation section.

% Neuroscientific investigations into cortical morphological
% changes/differences have illuminated interesting correlations with
% normal and pathological neurodevelopment in  addition to cognitive function.
Historically rooted in the meticulous work of von Economo \citep{economo2008},
imaging-based structural analysis of the brain plays a fundamental role
in identifying the relationship between cortical morphology, disease and cognition.
Quantitative cortical measures have been demonstrated in conditional 
abnormalities such as 
Huntington's disease \citep{rosas2002,rosas2005,selemon2004}, 
schizophrenia \citep{nesvag2008}, bipolar disorder \cite{lyoo2006}, Alzheimer's disease and frontotemporal
dementia \citep{du2007,dickerson2009}, Parkinson's disease \citep{jubault2011}, Williams syndrome \cite{thompson2005},
multiple sclerosis \citep{ramasamy2009}, autism \citep{chung2005,hardan2006},
migraines \citep{dasilva2007}, chronic smoking \citep{kuhn2010}, alcoholism \citep{fortier2011},
cocaine addiction \citep{makris2008}, Tourette syndrome in children \citep{sowell2008},
scoliosis in
female adolescents \citep{wang2012}, obsessive compulsive
disorder \citep{shin2007}, ADHD \citep{almeida-montes2012}, obesity \citep{raji2010}, and heritable \citep{peterson2009}
and elderly \citep{ballmaier2004} depression.  Cortical thickness also
varies normally as a function of age \citep{kochunov2011},
gender \citep{luders2006a}, untreated
male-to-female transsexuality \cite{luders2012},  handedness
\citep{luders2006,amunts2007}, intelligence \citep{shaw2006}, athletic
ability \citep{wei2011}, musical ability \citep{bermudez2009,foster2010}, 
tendency toward criminality \citep{raine2011}, and Tetris-playing
ability in female adolescents \citep{haier2009}.  Additionally,
recent studies demonstrate possible functional 
connectivity relationships using cortical thickness measures
\cite{worsley2005,lerch2006,he2007}.
Although these findings
are subject to debate and interpretation \citep{gernsbacher2007}, 
the availability of quantitative
computational methods for extracting such information
has proven invaluable for developing and refining fundamental 
neuroscience hypotheses.

@@

In this work, we describe our Insight Toolkit (ITK)-based cortical thickness pipeline
which is freely available as part of the Advanced Normalization Tools
(ANTs) software package.  This includes all the necessary preprocessing steps consisting
of well-vetted previously published algorithms for bias correction \cite{tustison2010},
brain extraction \cite{avants2010a}, $n$-tissue segmentation \cite{avants2011a},
template construction \cite{avants2010}, and image normalization \cite{avants2011}.
We also describe improvements made to the original DiReCT algorithm \cite{das2009}.
Equally as important, we provide explicit coordination between
these pipeline components complete with a set of useful parameters
which are employed to analyze the publicly available IXI data set. The
full pipeline and parameter set is 
encapsulated in a well-documented shell script which is also available in ANTs. 
Furthermore, we provide all the derived image data and other processing scripts on 
the NITRC repository specifically meant for this publication.  The
availability of both the code and data permits
the set of results described in this work to be fully reproducible.  This
permits other researchers to contrast their own results against
this baseline processing and to adapt the given volumetric pipeline for measuring
cortical thickness with their own datasets.

>>>>

% Neuroscientific investigations into cortical morphological
% changes/differences have illuminated interesting correlations with
% normal and pathological neurodevelopment in  addition to cognitive function.
Historically rooted in the meticulous work of von Economo \citep{economo2008},
imaging-based structural analysis of the brain plays a fundamental role
in identifying the relationship between cortical morphology, disease and cognition.
Quantitative cortical measures have been demonstrated in conditional 
abnormalities such as 
Huntington's disease \citep{rosas2002,rosas2005,selemon2004}, 
schizophrenia \citep{nesvag2008}, bipolar disorder \cite{lyoo2006}, Alzheimer's disease and frontotemporal
dementia \citep{du2007,dickerson2009}, Parkinson's disease \citep{jubault2011}, Williams syndrome \cite{thompson2005},
multiple sclerosis \citep{ramasamy2009}, autism \citep{chung2005,hardan2006},
migraines \citep{dasilva2007}, chronic smoking \citep{kuhn2010}, alcoholism \citep{fortier2011},
cocaine addiction \citep{makris2008}, Tourette syndrome in children \citep{sowell2008},
scoliosis in
female adolescents \citep{wang2012}, obsessive compulsive
disorder \citep{shin2007}, ADHD \citep{almeida-montes2012}, obesity \citep{raji2010}, and heritable \citep{peterson2009}
and elderly \citep{ballmaier2004} depression.  Cortical thickness also
varies normally as a function of age \citep{kochunov2011},
gender \citep{luders2006a}, untreated
male-to-female transsexuality \cite{luders2012},  handedness
\citep{luders2006,amunts2007}, intelligence \citep{shaw2006}, athletic
ability \citep{wei2011}, musical ability \citep{bermudez2009,foster2010}, 
tendency toward criminality \citep{raine2011}, and Tetris-playing
ability in female adolescents \citep{haier2009}.  Additionally,
recent studies demonstrate possible functional 
connectivity relationships using cortical thickness measures
\citep{worsley2005,lerch2006,he2007,chen2008}.
Although all these findings
are subject to debate and interpretation \citep{gernsbacher2007}, 
the availability of quantitative
computational methods for extracting such information
has proven invaluable for developing and refining fundamental 
neuroscience hypotheses.

@@

In this work, we describe our Insight Toolkit (ITK)-based cortical thickness pipeline
which is freely available as part of the Advanced Normalization Tools
(ANTs) software package.  This includes all the necessary preprocessing steps consisting
of well-vetted previously published algorithms for bias correction \cite{tustison2010},
brain extraction \cite{avants2010a}, $n$-tissue segmentation \cite{avants2011a},
template construction \cite{avants2010}, and image normalization \cite{avants2011}.
We also describe improvements made to the original DiReCT algorithm \cite{das2009}.
Equally as important, we provide explicit coordination between
these pipeline components within a set of well-documented shell scripts which 
is also available in the ANTs repository where parameters have been tuned
by the original authors and provide good performance across a number of data sets.
Furthermore, we provide all the derived image data and other processing scripts on 
the NITRC repository specifically meant for this publication.  The
availability of both the code and data permits
the set of results described in this work to be fully reproducible.  This
permits other researchers to contrast their own results against
this baseline processing and to adapt the given volumetric pipeline for measuring
cortical thickness with their own datasets.

### 2013-07-07 KapowskiChronicles/methods.tex
ENH:  Adding evaluation section.

The ANTs-based cortical thickness estimation workflow is illustrated 
in Figure \ref{fig:pipeline}.  The steps are as follows:
\begin{enumerate}
  \item initial N4 bias correction on input anatomical MRI,
  \item brain extraction using a hybrid segmentation/template-based strategy,
  \item alternation between prior-based segmentation and white matter posterior
        probability weighted bias correction,
  \item DiReCT-based cortical thickness estimation, and
  \item optional normalization to specified template.
\end{enumerate}
Each component, including both software and data, is briefly detailed 
below with the relevant references for additional information.

%We also note that each component is publicly available with all ANTs 
%algorithms available as open source.%
%\footnote{
%http://www.picsl.upenn.edu/ANTS
%}
Additionally, the coordination of all the algorithmic components is
encapsulated in the shell scripts \verb#antsCorticalThickness.sh# with
subcomponents delegated to \verb#antsBrainExtraction.sh# 
and \verb#antsAtroposN4.sh#.  This includes
optimal parameters for each of the algorithmic components which has worked
well for our processing and which are used to acquire the results 
described in this work.

@@

\begin{figure*}
  \centering
  \includegraphics[width=180mm]{Figures/Kapowski_pipeline2.pdf}
  \caption{Illustration of the main components of the ANTs processing 
  workflow containing all elements for determining cortical thickness. 
  We also included the domain of operations for select scripts also available
  in ANTs.
  Not shown is the optional subject to template registration.}
  \label{fig:pipeline}
\end{figure*}

@@

The ANTs implementation of this technique is currently available as a shell script, 
\verb#buildtemplateparallel.sh# (a multivariate version is also available as
\verb#antsMultivariateTemplateConstruction.sh#).  Both scripts are distributed as part of
 the ANTs repository.
The latter script permits the construction of multimodal templates (e.g. 
T1-weighted, T2-weighted, proton density MRI and fractional anisotropy).  
Both scripts accommodate a variety of computational resources
for facilitating template construction.  These computational resource possibilities include:
\begin{itemize}
  \item serial processing on a single workstation, 
  \item parallelized processing on a single workstation with multiple cores using \verb#pexec#%
  \footnote{http://www.gnu.org/software/pexec/pexec.1.html},
  \item parallelized processing using Apple's XGrid technology%
  \footnote{https://developer.apple.com/hardwaredrivers/hpc/xgrid\_intro.html}, 
  \item parallelized processing using Sun Grid Engine for cluster-based systems%
  \footnote{http://www.oracle.com/technetwork/oem/grid-engine-166852.html}, and 
  \item parallelized processing using the Portable Batch System for cluster-based systems%
  \footnote{http://www.pbsworks.com/}.
\end{itemize}

@@

Due to the important interplay between segmentation and bias correction,
we perform multiple N4 $\Leftrightarrow$ Atropos iterations.
In order to better integrate Atropos and N4, we use  
a pure tissue probability weight mask generated from the 
posterior probabilities generated from the segmentation 
process.  Given $N$ labels and the corresponding $N$
posterior probability maps $\{ P_1, \ldots, P_N\}$ produced
during the segmentation process, the N4 weight mask is 
created at each N4 $\Leftrightarrow$ Atropos iteration from
\begin{align}
  P_{pure\,\,tissue}(\mathbf{x}) = \sum_{i=1}^N P_i(\mathbf{x}) \prod_{j=1, j \neq i}^N \left( 1 - P_j(\mathbf{x}) \right).
\end{align}
One of the key insights of the original N3 algorithm is the
observation that inhomogeneities cause the intensity values of
pure tissue peaks to spread in the intensity histogram as though
convolved with a Gaussian.  A core contribution of N3 is the
proposed corrective of deconvolving the intensity histogram to 
accentuate the tissue peaks coupled with a spatial smoothing 
constraint. The pure tissue probability mask
weights more heavily the voxels corresponding to pure tissue 
types (as determined by the segmentation) during the deconvolution process 
while minimizing the contribution of regions such as the gray/white matter 
interface where peak membership is ambiguous.

@@

DiReCT was introduced 
in \cite{das2009} and made available in ANTs with the program \verb#KellySlater#.
Since then several improvements have been made and incorporated into the program
\verb#KellyKapowski#.%
\footnote{
Traditional academic discourse encountered in the published literature
rarely contextualizes peculiarities such as algorithmic nomenclature.
We briefly mention that
this was the source of a rare disagreement between the first two authors
based, as many disagreements are, on a simple misunderstanding and not an
affronting existential statement concerning a certain favorite sitcom
of the author's youth. 
}
Among the most significant advancements is that the more recent
implementation is multi-threaded, written in rigorous ITK coding style, and 
has been made publicly available through  ANTs complete with a unique user 
interface design developed specifically for ANTs tools.

@@

\begin{table}
\centering
\begin{tabular*}{0.475\textwidth}{@{\extracolsep{\fill}} l l}
\toprule
  1) L occipital lobe & 2) R occipital lobe \\
  3) L cingulate gyrus & 4) R cingulate gyrus \\
  5) L insula gyrus & 6) R insula gyrus \\
  7) L temporal lobe & 8) R temporal lobe \\
  9) L superior temporal gyr. & 10) R superior temporal gyr. \\
  11) L infero temporal region & 12) R infero temporal region \\
  13) L parahippocampal gyr. & 14) R parahippocampal gyr. \\
  15) L frontal pole & 16) R frontal pole \\
  17) L superior frontal gyrus & 18) R superior frontal gyrus \\
  19) L middle frontal gyrus & 20) R middle frontal gyrus \\
  21) L inferior gyrus & 22) R inferior gyrus \\
  23) L orbital frontal gyrus & 24) R orbital frontal gyrus \\
  25) L precentral gyrus & 26) R precentral gyrus \\
  27) L superior parietal lobule & 28) R superior parietal lobule \\
  29) L inferior parietal lobule & 30) R inferior parietal lobule \\
  31) L postcentral gyrus & 32)   R postcentral gyrus \\  
\bottomrule
\end{tabular*}
\caption{The 32 cortical NIREP labels.}
\label{table:nirep_labels}
\end{table}

@@

\paragraph{IXI}
The IXI data%
\footnote{
http://biomedic.doc.ic.ac.uk/brain-development/
}
consists of 561 healthy subjects (312 females, 249 males) imaged at three sites 
with several modalities acquired (T1-weighted, T2-weighted, proton density, magnetic 
resonance angiography, and diffusion tensor imaging).  The 
database also consists of  demographic information such as date of birth, date
of scan, weight,
height, ethnicity, occupation category, educational level, and marital status.
Several subjects were removed during the analysis since certain
demographic information was missing preventing an accurate estimate of 
the age at the time of image acquisition.

\paragraph{Oasis}
The Open Access Series of Imaging Studies (OASIS)%
\footnote{
http://www.oasis-brains.org/
}
 project consists of 416
(160 males, 256 females) T1-weighted images spanning 
the ages between 18 and 96 with 100 of these subjects having been 
diagnosed with very mild to moderate Alzheimer's disease.  
All subjects are right-handed.  Additional processing is included
with the distribution including bias corrected and segmented images
(gray, white and CSF matters).

>>>>

The ANTs-based cortical thickness estimation workflow is illustrated 
in Figure \ref{fig:pipeline}.  The steps are as follows:
\begin{enumerate}
  \item initial N4 bias correction on input anatomical MRI,
  \item brain extraction using a hybrid segmentation/template-based strategy,
  \item alternation between prior-based segmentation and ``pure tissue'' 
        posterior probability weighted bias correction,
  \item DiReCT-based cortical thickness estimation, and
  \item optional normalization to specified template.
\end{enumerate}
Each component, including both software and data, is briefly detailed 
below with the relevant references for additional information.

%We also note that each component is publicly available with all ANTs 
%algorithms available as open source.%
%\footnote{
%http://www.picsl.upenn.edu/ANTS
%}
The coordination of all the algorithmic components is
encapsulated in the shell scripts \verb#antsCorticalThickness.sh# with
subcomponents delegated to \verb#antsBrainExtraction.sh# 
and \verb#antsAtroposN4.sh#.  This includes
optimal parameters for each of the algorithmic components which has worked
well for our processing and are used to acquire the results 
described in this work.

@@

\begin{figure*}
  \centering
  \includegraphics[width=140mm]{Figures/Kapowski_pipeline2.pdf}
  \caption{Illustration of the main components of the ANTs processing 
  workflow containing all elements for determining cortical thickness. 
  We also included the domain of operations for the selected scripts.
  Not shown is the optional subject to template registration.}
  \label{fig:pipeline}
\end{figure*}

@@

The ANTs implementation of this technique is currently available as a shell script, 
\verb#buildtemplateparallel.sh# (a generalized, multivariate version is also available as
\verb#antsMultivariateTemplateConstruction.sh#).  Both scripts are distributed as part of
 the ANTs repository.
The latter script permits the construction of multimodal templates (e.g. 
T1-weighted, T2-weighted, proton density MRI and fractional anisotropy).  
Both scripts accommodate a variety of computational resources
for facilitating template construction.  These computational resource possibilities include:
\begin{itemize}
  \item serial processing on a single workstation, 
  \item parallelized processing on a single workstation with multiple cores using \verb#pexec#%
  \footnote{http://www.gnu.org/software/pexec/pexec.1.html},
  \item parallelized processing using Apple's XGrid technology%
  \footnote{https://developer.apple.com/hardwaredrivers/hpc/xgrid\_intro.html}, 
  \item parallelized processing using Sun Grid Engine for cluster-based systems%
  \footnote{http://www.oracle.com/technetwork/oem/grid-engine-166852.html}, and 
  \item parallelized processing using the Portable Batch System for cluster-based systems%
  \footnote{http://www.pbsworks.com/}.
\end{itemize}

@@

Due to the important interplay between segmentation and bias correction,
we perform multiple N4 $\rightleftharpoons$ Atropos iterations.
In order to better integrate Atropos and N4, we use  
a pure tissue probability weight mask generated from the 
posterior probabilities derived from the segmentation 
process.  Given $N$ labels and the corresponding $N$
posterior probability maps $\{ P_1, \ldots, P_N\}$ produced
during the segmentation process, the N4 weight mask is 
created at each N4 $\rightleftharpoons$ Atropos iteration from
\begin{align}
  P_{pure\,\,tissue}(\mathbf{x}) = \sum_{i=1}^N P_i(\mathbf{x}) \prod_{j=1, j \neq i}^N \left( 1 - P_j(\mathbf{x}) \right).
\end{align}
One of the key insights of the original N3 algorithm is the
observation that inhomogeneities cause the intensity values of
pure tissue peaks to spread in the intensity histogram as though
convolved with a Gaussian.  A core contribution of N3 is the
proposed corrective of deconvolving the intensity histogram to 
accentuate the tissue peaks coupled with a spatial smoothing 
constraint. The pure tissue probability mask
weights more heavily the voxels corresponding to pure tissue 
types (as determined by the segmentation) during the deconvolution process 
while minimizing the contribution of regions such as the gray/white matter 
interface where peak membership is ambiguous.

@@

DiReCT was introduced 
in \cite{das2009} and made available in ANTs as the program \verb#KellySlater#.
Since then several improvements have been made and incorporated into the program
\verb#KellyKapowski#.%
\footnote{
Traditional academic discourse encountered in the published literature
rarely contextualizes peculiarities such as algorithmic nomenclature.
We briefly mention that
this was the source of a rare disagreement between the first two authors
based, as many disagreements are, on a simple misunderstanding and not an
affronting existential statement concerning a certain favorite sitcom
of the author's youth. 
}
Among the most significant advancements is that the more recent
implementation is multi-threaded, written in rigorous ITK coding style, and 
has been made publicly available through  ANTs complete with a unique user 
interface design developed specifically for ANTs tools.

@@

\begin{table}
\centering
\begin{tabular*}{0.9\textwidth}{@{\extracolsep{\fill}} l l}
\toprule
  1) L occipital lobe & 2) R occipital lobe \\
  3) L cingulate gyrus & 4) R cingulate gyrus \\
  5) L insula gyrus & 6) R insula gyrus \\
  7) L temporal lobe & 8) R temporal lobe \\
  9) L superior temporal gyr. & 10) R superior temporal gyr. \\
  11) L infero temporal region & 12) R infero temporal region \\
  13) L parahippocampal gyr. & 14) R parahippocampal gyr. \\
  15) L frontal pole & 16) R frontal pole \\
  17) L superior frontal gyrus & 18) R superior frontal gyrus \\
  19) L middle frontal gyrus & 20) R middle frontal gyrus \\
  21) L inferior gyrus & 22) R inferior gyrus \\
  23) L orbital frontal gyrus & 24) R orbital frontal gyrus \\
  25) L precentral gyrus & 26) R precentral gyrus \\
  27) L superior parietal lobule & 28) R superior parietal lobule \\
  29) L inferior parietal lobule & 30) R inferior parietal lobule \\
  31) L postcentral gyrus & 32)   R postcentral gyrus \\  
\bottomrule
\end{tabular*}
\caption{The 32 cortical NIREP labels.}
\label{table:nirep_labels}
\end{table}

@@

\paragraph{IXI}
The IXI data%
\footnote{
http://biomedic.doc.ic.ac.uk/brain-development/
}
consists of 561 healthy subjects (312 females, 249 males) imaged at three sites 
with several modalities acquired (T1-weighted, T2-weighted, proton density, magnetic 
resonance angiography, and diffusion tensor imaging).  The 
database also consists of  demographic information such as date of birth, date
of scan, weight,
height, ethnicity, occupation category, educational level, and marital status.
Several subjects were removed after cortical thickness map calculation during 
the post-analysis 
since certain
demographic information was missing preventing an accurate estimate of 
the age at the time of image acquisition.

\paragraph{Oasis}
The Open Access Series of Imaging Studies (OASIS)%
\footnote{
http://www.oasis-brains.org/
}
 project consists of 416
(160 males, 256 females) T1-weighted images spanning 
the ages between 18 and 96 with 100 of these subjects having been 
diagnosed with very mild to moderate Alzheimer's disease.  
All subjects are right-handed.  Additional processing is included
with the distribution including bias corrected and segmented images
(gray, white and CSF matters).  20 subjects had repeat scans which
were used for evaluation of reproducibility.

### 2013-07-07 KapowskiChronicles/paper.tex
ENH:  Added thicknessplots.
ENH:  Adding conclusion section.
ENH:  Adding evaluation section.

%\documentclass[preprint,authoryear,review,12pt]{elsarticle}
\documentclass[final,5p,times,twocolumn]{elsarticle}

@@

\begin{abstract} 
Numerous studies have explored the relationship between cortical
structure and brain development, cognitive 
function and functional connectivity.
The highly convoluted cortical topography makes manual measurements
arduous and often impractical given the population sizes necessary 
for sufficient statistical power.  Computational techniques 
have permitted large-scale studies as they
provide robust and reliable localized measurements characterizing
the cortex with little or no human intervention.  Particularly useful
to the neuroscience community are publicly available tools, such as the popular 
surface-based Freesurfer, which facilitate the testing and refinement of 
hypotheses.  In this paper, we introduce the volume-based 
Advanced Normalization Tools (ANTs) cortical thickness automated pipeline
comprising well-vetted components such as SyGN (multivariate template construction),
SyN (image registration), N4 (bias correction), Atropos ($n$-tissue segmentation), 
and DiReCT (cortical thickness) all developed as part of the ANTs open science effort.  
Complementing the open source aspect of ANTs we investigate several hypotheses
explored previously in the literature using four different publicly 
available data sets totaling approximately 1200 subjects.  
To further promote open science and use of the proposed tools,
all scripts and templates used to produce the results
are hosted on the Neuroimaging Informatics Tools and Resources Clearinghouse (NITRC) website.
\end{abstract}

@@

%\input{results}

\section{Discussion and Conclusions}

>>>>

\documentclass[preprint,authoryear,review,12pt]{elsarticle}
% \documentclass[final,5p,times,twocolumn]{elsarticle}

@@

\begin{abstract} 
Numerous studies have explored the relationship between cortical
structure and brain development, cognitive 
function and functional connectivity.
The highly convoluted cortical topography makes manual measurements
arduous and often impractical given the population sizes necessary 
for sufficient statistical power.  Computational techniques 
have permitted large-scale studies as they
provide robust and reliable localized measurements characterizing
the cortex with little or no human intervention.  Particularly useful
to the neuroscience community are publicly available tools, such as the popular 
surface-based Freesurfer, which facilitate the testing and refinement of 
hypotheses.  In this paper, we introduce the volume-based 
Advanced Normalization Tools (ANTs) cortical thickness automated pipeline
comprising well-vetted components such as SyGN (multivariate template construction),
SyN (image registration), N4 (bias correction), Atropos ($n$-tissue segmentation), 
and DiReCT (cortical thickness) all developed as part of the ANTs open science effort.  
Four open data sets (IXI, Kirby, NKI, and Oasis), comprising approximately 
1200 subjects, were processed using the proposed pipeline.  Reproducibility scores
from a subset of these data, along with a complementary age prediction
evaluation, demonstrate robust performance.  Additionally, we present novel
gender differences in predicting structural connectivity across the lifespan.
To further promote open science and use of the proposed tools,
all scripts and templates used to produce the results
are hosted on the Neuroimaging Informatics Tools and Resources Clearinghouse (NITRC) website.
\end{abstract}

@@

\input{results}

\input{conclusions}
